---
title: "MiSS Descriptive Statistics"
output:
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
---

## 0. Data Wragling and Inspection

```{r}
library(tidyverse)

df_summary <- read_csv("summary.csv")

df_languages <- read_tsv("Table_of_Languages.tab")

head(df_summary)
head(df_languages)
```

```{r}
df_summary %>%
  group_by(model) %>%
  summarise(n_lang = n_distinct(lang))
```

```{r}
df_summary <- df_summary %>%
  separate(lang, into = c("ISO_639", "Script"), sep = "_", remove = TRUE)

head(df_summary)
```

```{r}
df_merged <- df_summary %>%
  left_join(df_languages, by = "ISO_639")

head(df_merged)
```

```{r}
# Define the hierarchical order of EGIDS levels
egids_levels <- c("Institutional", "Developing", "Vigorous", "In_Trouble", "Dying", "Extinct")

# Create the categorical variable and convert it to an ordered factor
df_merged <- df_merged %>%
  mutate(
    # Map 0/1 binary indicator flags to corresponding string labels
    EGIDS_level = case_when(
      Institutional == 1 ~ "Institutional", # EGIDS level from 0 to 4
      Developing == 1    ~ "Developing",    # EGIDS level = 5
      Vigorous == 1      ~ "Vigorous",      # EGIDS level = 6a
      In_Trouble == 1    ~ "In_Trouble",    # EGIDS level = 6b or 7
      Dying == 1         ~ "Dying",         # EGIDS level = 8a to 9
      Extinct == 1       ~ "Extinct",       # EGIDS level = 10
      TRUE               ~ NA_character_
    ),
    # Convert EGIDS_level to an ordered factor using the pre-defined levels
    EGIDS_level = factor(EGIDS_level, levels = egids_levels, ordered = TRUE)
  )

# Verify the creation and inspect frequency counts (including any NAs)
table(df_merged$EGIDS_level, useNA = "ifany")
```

```{r}
sum(is.na(df_merged$EGIDS)) # equals to the NAs in the table above
```

```{r}
dls_levels <- c("Still", "Emerging", "Ascending", "Vital","Thriving")

df_merged <- df_merged %>%
  mutate(Dls_Level = factor(Dls_Level, levels = dls_levels, ordered = TRUE))

# Verify the underlying order structure and inspect level counts
levels(df_merged$Dls_Level)
table(df_merged$Dls_Level, useNA = "ifany")
```

## 0.5 [Option] Filter only Mistral Small 3.2

```{r}
# make the following codes into comment if you want to analyze all models instead of only Mistral Small 3.2
df_merged <- df_merged %>%
  filter(model == "mistral-small-3.2")
```

## 1. Overall Statistics

```{r}
overall_stats <- df_merged %>%
  summarise(
    total_records = n(),
    unique_languages = n_distinct(ISO_639),
    mean_accuracy = mean(accuracy, na.rm = TRUE),
    sd_accuracy = sd(accuracy, na.rm = TRUE),
    median_accuracy = median(accuracy, na.rm = TRUE),
    iqr_accuracy = IQR(accuracy, na.rm = TRUE),
    min_accuracy = min(accuracy, na.rm = TRUE),
    max_accuracy = max(accuracy, na.rm = TRUE)
  )
print(overall_stats)
```

## 2. Accuracy by Model and Task

```{r}
model_task_stats <- df_merged %>%
  group_by(model, task) %>%
  summarise(
    n_obs = n(),
    mean_acc = mean(accuracy, na.rm = TRUE),
    median_acc = median(accuracy, na.rm = TRUE),
    sd_acc = sd(accuracy, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(model, desc(mean_acc))
print(model_task_stats)
```

```{r}
# Reshape to Model x Task Mean Accuracy Matrix
model_task_matrix <- df_merged %>%
  group_by(model, task) %>%
  summarise(mean_acc = mean(accuracy, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = task, 
    values_from = mean_acc
  )

print(model_task_matrix)
```

## 3. Language Vitality Analysis

```{r}
# Accuracy across different EGIDS levels
egids_stats <- df_merged %>%
  group_by(EGIDS_level) %>%
  summarise(
    n_languages = n_distinct(ISO_639),
    n_obs = n(),
    mean_acc = mean(accuracy, na.rm = TRUE),
    median_acc = median(accuracy, na.rm = TRUE),
    sd_acc = sd(accuracy, na.rm = TRUE),
    .groups = "drop"
  )
print(egids_stats)
```

```{r}
# Accuracy across different EGIDS scores (EGIDS level = 10 = Extinct)
egids_stats2 <- df_merged %>%
  group_by(EGIDS) %>%
  summarise(
    n_languages = n_distinct(ISO_639),
    n_obs = n(),
    mean_acc = mean(accuracy, na.rm = TRUE),
    median_acc = median(accuracy, na.rm = TRUE),
    sd_acc = sd(accuracy, na.rm = TRUE),
    .groups = "drop"
  )
print(egids_stats2)
```

## 4. Digital Language Support Aanlysis

```{r}
# Accuracy across different DLS levels
dls_stats <- df_merged %>%
  group_by(Dls_Level) %>%
  summarise(
    n_languages = n_distinct(ISO_639),
    n_obs = n(),
    mean_acc = mean(accuracy, na.rm = TRUE),
    median_acc = median(accuracy, na.rm = TRUE),
    sd_acc = sd(accuracy, na.rm = TRUE),
    .groups = "drop"
  )
print(dls_stats)
```

## 5. Classification (Family & Script) Analysis

```{r}
# Accuracy by Language Family (filtering for families with >= 5 languages)
family_stats <- df_merged %>%
  group_by(Family) %>%
  summarise(
    n_languages = n_distinct(ISO_639),
    n_obs = n(),
    mean_acc = mean(accuracy, na.rm = TRUE),
    median_acc = median(accuracy, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_languages >= 5) %>%
  arrange(desc(mean_acc))
print(family_stats)
```

```{r}
script_stats <- df_merged %>%
  group_by(Script) %>%
  summarise(
    n_languages = n_distinct(ISO_639),
    mean_acc = mean(accuracy, na.rm = TRUE),
    median_acc = median(accuracy, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n_languages))
print(script_stats)
```

## 6. Accuracy by L1 speakers

```{r}
# Speaker counts span multiple orders of magnitude; calculate log10 scale
df_merged <- df_merged %>%
  mutate(log10_L1 = log10(L1_Users + 1))

# Spearman & Pearson correlation between L1 speaker counts and accuracy
l1_correlation <- df_merged %>%
  filter(!is.na(L1_Users)) %>%
  summarise(
    spearman_corr = cor(log10_L1, accuracy, method = "spearman"),
    pearson_corr = cor(log10_L1, accuracy, method = "pearson")
  )
print(l1_correlation)

# Group L1 Users into ordinal bins for tabular presentation
df_merged <- df_merged %>%
  mutate(
    L1_Bin = case_when(
      L1_Users == 0 ~ "0 (Extinct)",
      L1_Users < 1e4 ~ "< 10K",
      L1_Users < 1e6 ~ "10K - 1M",
      L1_Users < 1e7 ~ "1M - 10M",
      TRUE ~ "> 10M"
    ),
    L1_Bin = factor(L1_Bin, levels = c("0 (Extinct)", "< 10K", "10K - 1M", "1M - 10M", "> 10M"))
  )

l1_bin_stats <- df_merged %>%
  group_by(L1_Bin) %>%
  summarise(
    n_languages = n_distinct(ISO_639),
    mean_acc = mean(accuracy, na.rm = TRUE),
    median_acc = median(accuracy, na.rm = TRUE),
    .groups = "drop"
  )
print(l1_bin_stats)
```

## 7. Interaction Analysis (Model x EGIDS_level)

```{r}
model_egids_interaction <- df_merged %>%
  group_by(model, EGIDS_level) %>%
  summarise(
    mean_acc = mean(accuracy, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = EGIDS_level, values_from = mean_acc)
print(model_egids_interaction)
```

## 8. Accuracy by Language x Script x Model

```{r}
# Calculate mean accuracy for each Language x Script x Model combination
combo_acc <- df_merged %>%
  group_by(Language_Name, ISO_639, Script, model) %>%
  summarise(mean_accuracy = mean(accuracy, na.rm = TRUE), .groups = "drop")

# Top 10 Highest Accuracy Combinations
top_10 <- combo_acc %>%
  slice_max(mean_accuracy, n = 10, with_ties = FALSE)

# Bottom 10 Lowest Accuracy Combinations
bottom_10 <- combo_acc %>%
  slice_min(mean_accuracy, n = 10, with_ties = FALSE)

print(top_10)
print(bottom_10)
```

## 9. A Look into some Major Languages (within one model)

```{r}
# Define target ISO language codes
target_langs <- c("eng", "cmn", "fra", "spa")

task_lang_acc <- df_merged %>%
  # Filter for target model and target languages
  filter(model == "mistral-small-3.2", ISO_639 %in% target_langs) %>%
  group_by(task, ISO_639) %>%
  summarise(
    mean_accuracy = mean(accuracy, na.rm = TRUE), 
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = ISO_639,
    values_from = mean_accuracy
  )

print(task_lang_acc)
```

## 10. Visulization

```{r}
library(ggplot2)

egids_model_trend <- df_merged %>%
  filter(!is.na(EGIDS_level)) %>%
  group_by(model, EGIDS_level) %>%
  summarise(mean_acc = mean(accuracy, na.rm = TRUE), .groups = "drop")

ggplot(egids_model_trend, aes(x = EGIDS_level, y = mean_acc, color = model, group = model)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_text(aes(label = sprintf("%.3f", mean_acc)), vjust = -0.8, size = 3.5, show.legend = FALSE) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(limits = c(0.4, 1.0), expand = expansion(mult = c(0.05, 0.15))) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 10, face = "bold"),
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    title = "Model Performance Trajectory Across EGIDS Levels",
    subtitle = "Interaction analysis: Model vs Language Vitality Level",
    x = "EGIDS Level",
    y = "Mean Accuracy",
    color = "Model"
  )
```

```{r}

# Heatmap: Model vs Task Mean Accuracy
ggplot(model_task_stats, aes(x = task, y = model, fill = mean_acc)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", mean_acc)), color = "black", size = 3.5, fontface = "bold") +
  scale_fill_gradient2(low = "#d73027", mid = "#fee08b", high = "#1a9850", midpoint = 0.7, name = "Mean Acc") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    title = "Model Accuracy Across LLM Tasks",
    x = "Task",
    y = "Model"
  )
```

```{r}
# Filter scripts with >= 2 languages
script_plot_data <- script_stats %>%
  filter(n_languages >= 2) %>%
  mutate(Script = reorder(Script, mean_acc))

ggplot(script_plot_data, aes(x = mean_acc, y = Script, fill = mean_acc)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.2f (n=%d)", mean_acc, n_languages)), hjust = -0.1, size = 3.5) +
  scale_fill_viridis_c(option = "magma", direction = -1) +
  scale_x_continuous(limits = c(0, 1.0), expand = expansion(mult = c(0, 0.15))) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14)) +
  labs(
    title = "Accuracy Disparity Across Writing Scripts",
    subtitle = "Scripts with at least 2 evaluated languages",
    x = "Mean Accuracy",
    y = "Script Code"
  )
```

```{r}
# Boxplot: Dls_Level vs Accuracy faceted by Model
df_merged %>%
  filter(!is.na(Dls_Level)) %>%
  ggplot(aes(x = Dls_Level, y = accuracy, fill = Dls_Level)) +
  geom_boxplot(outlier.alpha = 0.2, alpha = 0.7) +
  facet_wrap(~ model) +
  scale_fill_brewer(palette = "Spectral") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    title = "Accuracy Distribution Across Digital Language Support Levels",
    x = "Digital Language Support (Dls Level)",
    y = "Accuracy"
  )
```

## 11. English Error Analysis: 5 Random Mistakes per Task

Reading the per-item logs for `eng_Latn` / `mistral-small-3.2` (one file per task in `eng/`) to inspect where the model actually fails.

```{r}
task_order <- c("spelling", "inverse_spelling", "contains_character", "contains_word",
                "orthographic_similarity", "character_insertion", "word_insertion",
                "character_deletion", "word_deletion", "character_substitution",
                "word_substitution", "character_swapping", "word_swapping")

logs <- list.files("eng", pattern = "result_eng_Latn_.*_mistral-small-3\\.2\\.csv$",
                   full.names = TRUE) |> 
  map_dfr(~ read_csv(.x, show_col_types = FALSE))

set.seed(42)
mistakes <- logs |> 
  # real mistakes, not API errors
  filter(is_correct == 0, is_api_error == 0) |>
  group_by(task) |> 
  slice_sample(n = 5) |> 
  ungroup() |> 
  mutate(task = factor(task, levels = task_order)) |> 
  arrange(task) |> 
  select(task, input = input_key, expected, model_output)

write_csv(mistakes, "english_mistral_mistakes.csv")
print(mistakes, n = nrow(mistakes))
```

## 12. English Error Diagnosis: model mistake vs. scoring mistake

The prompts sent to the model are not stored in the log files, only the task parameters in `input_key`. So each task's rule is worked out from the `input_key` -\> `expected` pairs. One alternative reading of the rules (counting positions from 0) is tested against what the model actually wrote in 12.5.

```{r}
library(tidyverse); library(jsonlite)

logs <- list.files("eng", pattern = "result_eng_Latn_.*_mistral-small-3\\.2\\.csv$",
                   full.names = TRUE) |>
  map_dfr(~ read_csv(.x, show_col_types = FALSE)) |>
  filter(is_api_error == 0)          # drop API errors

# strip one or more layers of wrapping quotes. is_correct already tolerates straight double
# quotes, but not curly or single quotes, so this helper is more forgiving than is_correct
strip_q <- function(s) {
  s <- str_trim(replace_na(s, ""))
  repeat {
    n <- nchar(s)
    k <- which(n >= 2 &
               str_sub(s, 1, 1) %in% c('"', '“', '‘', "'") &
               str_sub(s, n, n) %in% c('"', '”', '’', "'"))
    if (!length(k)) break
    s[k] <- str_trim(str_sub(s[k], 2, nchar(s[k]) - 1))
  }
  s
}
# punctuation/case-insensitive comparison
loose <- function(s) {
  strip_q(s) |> str_replace_all("[“”]", '"') |>
    str_replace_all("[‘’]", "'") |>
    str_replace_all("[^\\w\\s]", "") |> str_squish() |> str_to_lower()
}
# helpers to pull fields out of the input_key JSON
kv <- function(x, f) map_chr(x, ~{v <- fromJSON(.x)[[f]]; if (is.null(v)) NA_character_ else as.character(v)})
kn <- function(x, f) map_dbl(x, ~{v <- fromJSON(.x)[[f]]; if (is.null(v)) NA_real_  else as.numeric(v)})
```

### 12.1 How much of the gap is scoring rather than the model?

Both comparisons are applied only to the four `word_` tasks. In the other tasks spaces and punctuation are part of the answer itself (e.g. `spelling` must output the spaces, and a word like `I'm` contains an apostrophe), so ignoring them would count real mistakes as correct.

```{r}
scoring_gap <- logs |>
  group_by(task) |>
  summarise(
    n                = n(),
    official_acc     = mean(is_correct),
    if_ws_ignored    = if (!str_starts(first(task), "word_")) NA_real_
                       else mean(str_squish(strip_q(model_output)) == str_squish(strip_q(expected))),
    if_punct_ignored = if (!str_starts(first(task), "word_")) NA_real_
                       else mean(loose(model_output) == loose(expected)),
    .groups = "drop"
  ) |>
  mutate(recoverable_pp = round(100 * (if_punct_ignored - official_acc), 1)) |>
  arrange(desc(recoverable_pp))
print(scoring_gap, n = Inf)
```

### 12.2 The double-space issue in `expected` (word tasks)

The story texts were flattened from several lines into one, leaving two spaces wherever a line break used to be ("The Hungry Frog␣␣“Take a break…"). The `expected` answers collapse these to one space, so when Mistral copies the spacing of its input it is scored wrong. Not every double space in (a) comes from the input, though: in `word_deletion` Mistral sometimes leaves two spaces where it removed a word ("out from␣␣the bushes" after deleting "behind"). Three checks below.

```{r}
# (a) rows scored wrong that become identical to `expected` once spacing is squeezed
ws_only <- logs |>
  filter(str_starts(task, "word_"), is_correct == 0,
         str_squish(strip_q(model_output)) == str_squish(strip_q(expected)),
         strip_q(model_output) != strip_q(expected))   # exclude quote-only diffs

ws_only |> count(task, name = "scored_wrong_on_spacing_alone") |> print()
cat("total:", nrow(ws_only), "rows\n")

ws_only |> slice_head(n = 2) |>
  transmute(task, expected = strip_q(expected), model_output = strip_q(model_output)) |>
  print()
```

```{r}
# (b) the double space never appears in expected, but often appears in model_output
logs |>
  filter(str_starts(task, "word_")) |>
  group_by(task) |>
  summarise(expected_with_double_space = sum(str_detect(expected, "  ")),
            model_output_with_double_space = sum(str_detect(strip_q(model_output), "  ")),
            .groups = "drop") |>
  print()
```

```{r}
# (c) input_len proves the extra space was in the INPUT, not invented by the model.
#     `input_len` is the character count of the passage actually sent to the model.
#     For word_substitution where nchar(old_word) == nchar(new_word) the edit preserves
#     length, so a faithful expected answer must satisfy  input_len == nchar(expected).
logs |>
  filter(task == "word_substitution",
         nchar(kv(input_key, "old_word")) == nchar(kv(input_key, "new_word"))) |>
  mutate(model_has_double_space = str_detect(strip_q(model_output), "  "),
         len_gap = input_len - nchar(expected)) |>
  count(model_has_double_space, len_gap) |>
  print()
```

Reading (c): when the model emits no double space the gap is 0 (392 rows) - `expected` matches the passage length. Whenever the model does emit a double space the gap is never 0, but 1 or 2 - `expected` is exactly as many characters short as the spaces that were collapsed. The 55 rows with gap 1 and no double space are cases where the model collapsed the spacing too.

### 12.3 Word length matters

A defective prompt would fail uniformly. Accuracy instead decays monotonically with word length in every character task — `character_deletion` is 99.6% on 2-3 letter words and 44% on 9+ letter words under the identical prompt.

```{r}
logs |>
  filter(str_starts(task, "character_")) |>
  mutate(len = nchar(kv(input_key, "word")),
         bin = cut(len, c(0, 3, 5, 8, Inf), labels = c("2-3", "4-5", "6-8", "9+"))) |>
  group_by(task, bin) |>
  summarise(acc = mean(is_correct), .groups = "drop") |>
  pivot_wider(names_from = bin, values_from = acc, id_cols = task) |>
  print()
```

### 12.4 Position

`character_insertion` is 8x more accurate when the anchor is the **last** character than when it is in the middle, i.e. exactly when "append to the end" happens to be the right answer.

```{r}
logs |>
  filter(task == "character_insertion") |>
  mutate(w = kv(input_key, "word"), a = kv(input_key, "after_char"),
         pos = map2_int(w, a, ~{i <- str_locate(.x, fixed(.y))[, 1]; if (is.na(i)) NA_integer_ else as.integer(i)})) |>
  filter(!is.na(pos)) |>
  mutate(anchor = case_when(pos == nchar(w) ~ "LAST  (append == correct)",
                            pos == 1        ~ "FIRST",
                            TRUE            ~ "MEDIAL")) |>
  group_by(anchor) |>
  summarise(acc = mean(is_correct), n = n(), .groups = "drop") |>
  print()

# and the corresponding failure signature: the char lands at the end of the word
logs |>
  filter(task == "character_insertion", is_correct == 0) |>
  mutate(appended = strip_q(model_output) == paste0(kv(input_key, "word"),
                                                    kv(input_key, "insert_char"))) |>
  summarise(errors = n(), appended_at_end = sum(appended),
            pct = 100 * mean(appended)) |> print()
```

### 12.5 Is the model counting positions from 0 instead of 1?

If the instruction never said whether the first letter is position 1 or position 0, the wrong answers would cluster on the "count from 0" version. Only 95 of 1,881 do, so that is not the explanation.

```{r}
swap_chr <- function(w, i, j) { v <- str_split_1(w, ""); tmp <- v[i]; v[i] <- v[j]; v[j] <- tmp; paste0(v, collapse = "") }

logs |>
  filter(task == "character_swapping", is_correct == 0) |>
  mutate(w = kv(input_key, "word"), p1 = kn(input_key, "pos1"), p2 = kn(input_key, "pos2"),
         zero_indexed = map_chr(seq_len(n()), ~{
           if (max(p1[.x], p2[.x]) + 1 > nchar(w[.x])) NA_character_
           else swap_chr(w[.x], p1[.x] + 1, p2[.x] + 1)})) |>
  summarise(errors = n(),
            matches_0_indexed = sum(strip_q(model_output) == zero_indexed, na.rm = TRUE),
            unchanged         = sum(strip_q(model_output) == w),
            same_letters_wrong_order = sum(map2_lgl(strip_q(model_output), expected,
                                    ~identical(sort(str_split_1(.x, "")), sort(str_split_1(.y, "")))))) |>
  print()
```

`same_letters_wrong_order` counts answers holding exactly the right characters in the wrong order: the model retrieves the letters but cannot place them. This count includes the `unchanged` rows (the original word also has the right letters), so the two columns overlap.

### 12.6 Annotated mistake sample

`whitespace only` and `punctuation only` apply to the `word_` tasks only (see 12.1). They are not labelled as scoring mistakes, because the difference can come from either side: most double spaces are copied from the input, but some are left by the model where it deleted a word (see 12.2).

```{r}
set.seed(42)
diagnosed <- logs |>
  filter(is_correct == 0) |>
  mutate(m = strip_q(model_output), e = strip_q(expected),
         word_task = str_starts(task, "word_"),
         # order matters: loose() also ignores spaces, so the whitespace check must come first
         diagnosis = case_when(
    word_task & m != e & str_squish(m) == str_squish(e)                      ~ "whitespace only",
    word_task & loose(model_output) == loose(expected)                       ~ "punctuation only",
    m == kv(input_key, "word")                                               ~ "MODEL: no edit performed",
    TRUE                                                                     ~ "MODEL: wrong edit"
  )) |>
  group_by(task) |> slice_sample(n = 5) |> ungroup() |>
  select(task, input = input_key, expected, model_output, diagnosis)

write_csv(diagnosed, "english_mistral_mistakes_diagnosed.csv")
print(diagnosed, n = Inf)
```
